An enhancer code compendium of early human neuronal and neural crest development

### 2025-05-15 manuscript.tex
Changes troughout

By encoding clusters of transcription factor (TF) binding sites, genomic enhancers act as hubs for combinatorial TF
recruitment. 
Enhancer code studies have been limited to specific cell types in model organisms. 
Here we set out to create a compendium of enhancer codes for all cell types of the developing human neural tube and neural crest. 
To this end, we generated a single-cell multiome atlas of human neural tube organoids and of the head of a four weeks post conception human embryo. 
Using DNA sequence based deep learning - DeepNeuralTube - we identify 15 cell types that are shared between the organoid 
model and the human embryo and comprehensively analyse the enhancer code of each individual cell type. 
We confirm the paradigm that enhancers are encoded using heterotypic clusters of 4-5 TF binding sites in a nucleosome depleted region, using soft syntactic rules. 
We also provide counter examples to this rule, with enhancers encoded as homotypic clusters and by dimer motifs. 
Finally, we validate DeepNeuralTube using enhancer reporter assays and assess the role of human genetic variants on 
facial mesenchyme enhancers. 
Our neural tube enhancer compendium is available as a resource at: something.aertslab.org

@@

In this study, we address these limitations by building an enhancer code compendium of cell types and cell states generated by human neural tube organoids, 
a model system for early human neural and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021}.
We compare this compendium to that of the head of a four weeks post conception human embryo.
Through these comparisons we assess the enhancer code paradigm, 
and also provide counter examples specifically related to the enhancer code underlying anterior-posterior patterning.
\par

@@

We generated human neural tube organoids to model early human neural and neural crest development. 
To this end we embedded single cells from two induced pluripotent stem cell (iPSC) lines (Sigma and BJ1) and one embryonic stem cell line (ESC; H9) into separate synthetic poly-ethylene glycol (PEG) based extracellular matrices. The organoids were incubated for 11 days and we supplemented the medium with retinoic acid (RA) and Smoothened agonist (SAG) from day 3 until day 5 (\textbf{Fig. 1a}). 
These conditions were previously shown to generate multiple neural progenitor and
neural crest states\cite{abdelfattahActuationEnhancesPatterning2021}. 
To characterize the organoid-derived cell types to a greater level of detail at both transcriptomic as well as chromatin accessibility level we performed 10x single-cell multiomics (combined scATAC-seq and
scRNA-seq) 
at four time points of differentiation (day 4-5, day 6-7, day 8-9 and day 10-11; \textbf{Fig. 1a}), 
resulting in 22,862 high quality cells (\textbf{Fig. 1b}). 
Note that similar transcriptomic states were obtained from all three cell lines (\textbf{Fig. S1a-b}) demonstrating the robustness of the organoid system.\par

@@

Finally, to assess whether the organoid-derived cell types and states are also similar to the embryo at the level of chromatin accessibility
we identified sets of co-accessible regions in the organoid model using topic modeling 
and identified equivalent cells in the embryo where the same regions were also co-accessible.
We observed similar patterns of marker gene expression for the same cell state across the organoid and embryo (\textbf{Fig. 1i}) 
with an average Pearson correlation coefficient of 0.83 across matched states (\textbf{Fig. S1c-d}).\par

@@

To gain insight into the sequence characteristics underlying cell type-specific chromatin accessibility of 
neural progenitor, neural crest and neural cell type-specific enhancers, 
we trained two sequence-to-function model convolutional neural networks:
"$DeepNeuralTube^o$" and "$DeepNeuralTube^e$", on the organoid and embryo atlases respectively. 
The models use the DNA sequence of genomic regions as input and classify regions based on their cell type- or state-specific chromatin accessibility (in a multi-label multi-class fashion; \textbf{Fig. 2a}). 
The classes of the models are defined as the topics from topic modeling. 
To get a fine-grained definition of cell states, we performed topic modeling separately for the subset of neural progenitors, neural crest (incl. facial mesenchyme and peripheral neurons) and early differentiating neurons for both organoid (to train $DeepNeuralTube^o$) and embryo (to train $DeepNeuralTube^e$).\par

@@

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) 
    \textbf{b}, receiver-operator curves for classifying neural tube (left) and facial mesenchyme enhancers using
    $DeepNeuralTube^o$ and $DeepNeuralTube^e$. 
    \textbf{c-e}, UMAP of TomTom\cite{guptaQuantifyingSimilarityMotifs2007} motif-to-motif similarities
    using both \textit{de novo} motifs from DeepNeuralTube and known enriched motifs. Each dot is a motif. Motifs are
    colored by: cluster identity (\textbf{c}), model of origin 
    (\textbf{d}; only showing \textit{de novo} motifs in this case) and whether the motif is found \textit{de novo} (Deep Learning) or by motif enrichment (cisTarget; \textbf{e}).
    \textbf{f}, Example representative motifs from clusters in b, showing motifs from motif enrichment (left) and found
    \textit{de novo} (right). 
    \textbf{g}, contribution score for reference and mutant version of VISTA enhancer hs111
    for $DeepNeuralTube^o$ (top two panels) and $DeepNeuralTube^e$ (bottom two panels) and enhancer-reporter
    assay for reference and mutant version of hs111 (right two panels)\cite{
    viselVISTAEnhancerBrowsera2007, kosickiVISTAEnhancerBrowser2025}. 
    \textbf{h}, contribution score for
    $DeepNeuralTube^o$ (top) and $DeepNeuralTube^e$ (bottom) of a genomic human neural crest
    enhancer (chr1:161200798-161201298; hg38) and enhancer-reporter assay in chicken embryo (HH4; right). Green is enhancer signal, pink is
    constitutive enhancer that serves as electroporation control.
    Also see Figure S2
  }
\end{figure}

@@

Focusing on neural progenitors, we identified topics representing cells and regions of different dorso-ventral
identities in both the organoid and the embryo data (\textbf{Fig. 3a}). 
From ventral to dorsal this is topic 36, 38, 33, 54 and 48 for the organoids and
topic 34, 38, 79, 88 and 58 for the embryo. 
As an example we first investigated two enhancers of \textit{SHH} that are strongly conserved and specific to the 
floorplate: SFPE1 and SFPE2\cite{jeongDistinctRegulatorsShh2003}. 
\textit{SHH} is most strongly expressed in the most ventral topic (33 and 34 resp. for the organoids and the embryo; \textbf{Fig. 3b}).
SFPE1 is located upstream of \textit{SHH} and SFPE2 in its second intron. Both enhancers are indeed most accessible in the cells belonging to the ventral topic (\textbf{Fig. 3c}). 
When we tested the enhancer activity of SFPE1 in chicken embryos we could confirm its floor plate specific activity
(\textbf{Fig. 3d}).
Explaining the accessibility of both enhancers using DeepNeuralTube reveals two FOX (one partial) and two RFX (one partial) binding sites in SFPE1 and two FOX, a ZEB and a TEAD binding site in SFPE2 (\textbf{Fig. 3e}).\par

To investigate the enhancer code underlying dorso-ventral patterning of neural progenitors further, we clustered 
and aligned \textit{de novo} motifs, identified in dorso-ventral regions, based on pairwise motif similarities. 
Next, we calculated the average contribution for all identified motif instances per cluster and for each topic 
resulting in a dorso-ventral enhancer code-table (\textbf{Fig. 3f}).
Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns underlying the dorso-ventral states and the 
importance of the instances of these patterns is highly correlated across both models (0.75 average Pearson correlation; \textbf{Fig. 3f}). 
For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD and RFX instances 
are important (\textbf{Fig. 3f-g}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo (\textbf{Fig. 3h} and Fig. S3). 
In subsequent more dorsal neural progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3f-h)}); 
PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3f-h} and Fig. S3) and 
ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3f-h} and Fig. S3). 
For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), 
SOX instances are important correlating with the expression of \textit{SOX2/3} (\textbf{Fig. 3f-h} and Fig. S3). 
Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3f} and Fig. S3). 
Finally, we calculated the Jaccard index of genomic regions based on the presence of instances of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig. 3i}). 
FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns tend to be present in separate regions (\textbf{Fig. 3i}).\par

@@

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) 
    \textbf{b}, Boxplots of the expression of \textit{SHH} in organoid (top) and embryo (bottom) across different dorso-  ventral topics ($n_{topic 33}$ = 4,571, $n_{topic 38}$ = 4,521, $n_{topic 36}$ = 6,120, $n_{topic 54}$ = 2,808, and $n_{topic 48}$ = 5,390 for organoid cells;
      and $n_{topic 34}$ = 637, $n_{topic 38}$ = 1,093, $n_{topic 79}$ = 3,939, $n_{topic 88}$ = 1,861, and $n_{topic 58}$ = 1,773 for embryo cells).  With n representing the number of cells.
    \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr7:155,799,664-155,827,483 ; hg38. SFPE2 and SFPE1 are highlighted. 
    \textbf{d}, Enhancer reporter assay in chicken embryo. Blue is DAPI, pink is constitutive electroportion control,
    green is GFP driven by SFPE1 enhancer.
    \textbf{e}, Nucleotide contribution scores for SFPE2 (left) and SFPE1 (right) for organoid topic 33 (top) and embryo topic 34 (bottom). 
    \textbf{f}, Code table showing average contribution score of pattern instances for each dorso-ventral topic. Contribution scores for $DeepNeuralTube^o$ are not outlined. Contribution scores for $DeepNeuralTube^e$ are outlined and negated. 
    \textbf{g}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) for dorso-ventral topics. 
    \textbf{h}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) dorso-ventral topics. 
    \textbf{i}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. 
    \textbf{j}, Nucleotide contribution score for chr18:77367159-77367659; hg38 for organoid topic 33 (top) and embryo
    topic 34 (bottom).
    \textbf{k}, Prediction scores for chr18:77367159-77367659; hg38 for doros-ventral topics in organoid (top) and
    embryo (bottom).
    \textbf{l}, Enhancer reporter assay in chicken embryo. Blue is DAPI, pink is constitutive electroportion control,
    green is GFP driven by chr18:77367159-77367659; hg38 enhancer. Scale bar represents \SI{500}{\micro\meter}. 
    Also see Figure S3 and S4.
}
\end{figure}

\subsection*{
  Homotypic clusters of distinct homeodomain transcription factor binding sites encode anterior-posterior patterning
}

@@

$DeepNeuralTube^e$ was able to distinguish the different AP domains but grouped topics of similar domains
% TODO: S.A.: How does the organoid model score all these AP topics?
together (i.e., Topic 59, 31 and 62; Topic 52 and 71), resulting in four distinct AP domains (\textbf{Fig.4 c-d}) and 
for this it made use of four distinct homeodomain TF binding motifs, one for each domain (\textbf{Fig. 4e}). 
From anterior to posterior this is \textit{TAATTA}, \textit{GGATTA}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA}.
Notably, the patterns corresponding to the two posterior most topics correspond to homeobox dimer-motifs. 
Previously it had been hypothesized that genome-wide specific binding of homeodomain TFs is accomplished through
cooperative binding of two TFs\cite{mannChapter3Hox2009}. 
Indeed, the presence of instances of the \textit{TCATGWTGANTGA} motif correlates with the expression of both \textit{PAX5/8} and \textit{EN1/2} 
and the prescence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig. 4d-e} and Fig. S5). 
In contrast, patterns corresponding to the two anterior most topics correspond to monomer motifs, 
correlating with the expression of \textit{LHX2/9}, \textit{RAX} and \textit{SIX3/6} for the most anterior topic and 
with the expression of \textit{OTX1/2}, \textit{EMX2}, \textit{DMBX1} and \textit{LMX1A} for the second most anterior
topic (\textbf{Fig. 4d-e} and Fig. S5)\par
%TODO: I'm surprised to see EMX2 (forebrain) together with OTX1/2 and DMBX1 (midbrain)

@@

\afterpage{\clearpage}
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/main/Figure_4.png}
  \caption{
    \textbf{An unsophisticated enhancer code underlies anterior-posterior patterning in neural progenitors.} \\
  \textbf{a}, UMAP of embryo neural progenitors colored by cell-topic probabilities in RGB color scales for
anterior-posterior topics. 
  \textbf{b}, Heatmap showing pairwise Pearson correlation coefficients of region-topic probabilities for dorso-ventral and anterior-posterior topics. 
  \textbf{c}, Heatmap showing prediction scores of $DeepNeuralTube^e$ on the topic 3,000 regions per topic (y-axis) for each anterior-posterior topics (x-axis). 
  \textbf{d}, Heatmap showing the log of the number of seqlets for each identified pattern (y-axis) for all anterior-posterior topic (x-axis). 
  \textbf{e}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in embryo
  anterior-posterior topics. 
  \textbf{f}, Boxplot showing prediction score before (wt) and after (alt) substituting all instances of \textit{CTAATTAG} with \textit{CTAATCCCT} (top) or substituting all instances of \textit{CTAATCCCT} with \textit{CTAATTAG} (bottom) in the topic 3,000 regions of either Topic 61 or Topic 31. 
  \textbf{g}, Boxplot showing prediction scores for all anterior-posterior topics after embedding 0 to 5 instances of \textit{CTAATTAG} while optimizing their location for Topic 61 (top) and after embedding 0 to 5 instances of \textit{CTAATCCCT} while optimzing their location for Topic 31 (bottom); n = 200. 
  \textbf{h}, Boxplot showing prediction scores for all anterior-posterior topics after embedding 0 to 5 instances of \textit{CTAATTAG} while optimizing their location for Topic 31 (top) and after embedding 0 to 5 instances of \textit{CTAATCCCT} while optimzing their location for Topic 61 (bottom); n = 200. 
  \textbf{i}, receiver-operator curve (ROC) (top) and precision-recall curve (bottom) for logistic regression classifier classifying genomic regions as either Topic 61 or Topic 31 based on motif scores for \textit{CTAATTAG} and \textit{CTAATCCCT} motifs. 
  U.: UMAP. 
  Also see Figure S5.
}
\end{figure}
\afterpage{\clearpage}

@@

\afterpage{\clearpage}
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/main/Figure_5.png}
  \caption{
    \textbf{Deciphering the enhancer code underlying pre-migratory and migratory neural crest and facial mesenchyme.} \\
  \textbf{a}, UMAP of organoid (top two) and embryo (bottom two) neural crest cells colored by cell-topic probabilities in RGB color scales for neural crest related topics. 
  \textbf{b}, Code table showing average contribution score of pattern instances for each neural crest related topic. Contribution scores for $DeepNeuralTube^o$ are not outlined. Contribution scores for $DeepNeuralTube^e$ are outlined and negated. 
  \textbf{c}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) of neural crest related topics. 
  \textbf{d}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) for neural crest related topics. 
  \textbf{e}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. 
  \textbf{f}, Manhattan plot showing -log p value of the SNP being associated with variations in facial shape\cite{whiteInsightsGeneticArchitecture2021} (positive y-axis) and the maximum (across classes of the model) of the absolute value of the difference in prediction score between the reference and alternative allele (negative y-axis; circles for $DeepNeuralTube^o$ and crosses for $DeepNeuralTube^e$) across the human genome. Lead SNPs are shown with a bigger dot size. 
  \textbf{g}, Zoom in of chromosome one with the locus of SNP rs1555067 shown as an inset, the SNP itself is shown in red. 
  \textbf{h}, Nucleotide contribution scores for Topic 58 of $DeepNeuralTube^o$ of the references (top) and alternative (bottom) allele of SNP rs1555067. The location of the SNP is indicated with a dotted line. 
  U: UMAP.
  Also see Figure S6 and S7.
  }
\end{figure}
\afterpage{\clearpage}

@@

During early neural differentiation, we identified several topics for different stages of neural fate
acquisition in both organoid and embryo cells (\textbf{Fig. 6a}). Again we made use of $DeepNeuralTube^o$ and
$DeepNeuralTube^e$ to generate a neuron code-table (\textbf{Fig. 6b}). 
Both models found similar patterns and the importance of the instances of these patterns is correlated across
both models (0.47 average Pearson correlation coefficient; \textbf{Fig. 6b}).\par

@@

\section*{MAIN FIGURE TITLES AND LEGENDS}

>>>>

By encoding clusters of transcription factor (TF) binding sites, genomic enhancers act as hubs for combinatorial TF
recruitment. 
Enhancer code studies have been limited to specific cell types in model organisms. 
Here we set out to create a compendium of enhancer codes for all cell types of the developing human neural tube and neural crest. 
To this end, we generated a single-cell multiome atlas of human neural tube organoids and of the head of a four weeks post conception human embryo. 
Using DNA sequence based deep learning - DeepNeuralTube - we identify 15 cell types that are shared between the organoid 
model and the human embryo and comprehensively analyse the enhancer code of each individual cell type. 
We confirm the paradigm that enhancers are encoded using heterotypic clusters of 4-5 TF binding sites in a nucleosome depleted region, using soft syntactic rules. 
We also provide counter examples to this rule, with enhancers encoded as homotypic clusters and by dimer motifs. 
Finally, we validate DeepNeuralTube using enhancer reporter assays and assess the role of human genetic variants on 
facial mesenchyme enhancers. 
We provide tools for building enhancer code compendia through our genomic deep learning package, CREsted.

@@

In this study, we address these limitations by building an enhancer code compendium of cell types and cell states generated by human neural tube organoids, 
a model system for early human neural and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021}.
We compare this compendium to that of the head of a four weeks post conception human embryo.
Through these comparisons we assess the enhancer code paradigm.
\par

@@

We generated human neural tube organoids to model early human neural and neural crest development. 
To this end we embedded single cells from two induced pluripotent stem cell (iPSC) lines (Sigma and BJ1) and one embryonic stem cell line (ESC; H9) into separate synthetic poly-ethylene glycol (PEG) based extracellular matrices. The organoids were incubated for 11 days and we supplemented the medium with retinoic acid (RA) and Smoothened agonist (SAG) from day 3 until day 5 (\textbf{Fig. 1a}). 
These conditions were previously shown to generate multiple neural progenitor and
neural crest states\cite{abdelfattahActuationEnhancesPatterning2021}. 
To characterize the organoid-derived cell types at both transcriptomic as well as chromatin accessibility level we performed 10x single-cell multiomics (combined scATAC-seq and scRNA-seq) 
at four time points of differentiation (day 4-5, day 6-7, day 8-9 and day 10-11; \textbf{Fig. 1a}), 
resulting in 22,862 high quality cells (\textbf{Fig. 1b}). 
Note that similar transcriptomic states were obtained from all three cell lines (\textbf{Fig. S1a-b}) demonstrating the robustness of the organoid system.\par

@@

Finally, to assess whether the organoid-derived cell types and states are also similar to the embryo at the level of chromatin accessibility
we identified sets of co-accessible regions in the organoid model using topic modeling 
and identified equivalent cells in the embryo (defined as cells where the same regions were also co-accessible).
We observed similar patterns of marker gene expression for the same cell state across the organoid and embryo (\textbf{Fig. 1i}) 
with an average Pearson correlation coefficient of 0.83 across matched states (\textbf{Fig. S1c-d}).\par

@@

To gain insight into the sequence characteristics underlying cell type-specific chromatin accessibility of 
neural progenitor, neural crest and neural cell type-specific enhancers, 
we trained two sequence-to-function model convolutional neural networks:
"$DeepNeuralTube^o$" and "$Deep$-$NeuralTube^e$", on the organoid and embryo atlases respectively. 
The models use the DNA sequence of genomic regions as input and classify regions based on their cell type- or state-specific chromatin accessibility (in a multi-label multi-class fashion; \textbf{Fig. 2a}). 
The classes of the models are defined as the topics from topic modeling. 
To get a fine-grained definition of cell states, we performed topic modeling separately for the subset of neural progenitors, neural crest (incl. facial mesenchyme and peripheral neurons) and early differentiating neurons for both organoid (to train $DeepNeuralTube^o$) and embryo (to train $DeepNeuralTube^e$).\par

@@

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) 
    \textbf{b}, receiver-operator curves for classifying neural tube (left) and facial mesenchyme enhancers using
    $DeepNeuralTube^o$ and $DeepNeuralTube^e$. 
    \textbf{c-e}, UMAP of TomTom\cite{guptaQuantifyingSimilarityMotifs2007} motif-to-motif similarities
    using both \textit{de novo} motifs from DeepNeuralTube and known enriched motifs. Each dot is a motif. Motifs are
    colored by: cluster identity (\textbf{c}), model of origin 
    (\textbf{d}; only showing known motifs in this case) and whether the motif is found \textit{de novo} (Deep Learning) or by motif enrichment (cisTarget; \textbf{e}).
    \textbf{f}, Example representative motifs from clusters in b, showing motifs from motif enrichment (left) and found
    \textit{de novo} (right). 
    \textbf{g}, contribution score for reference and mutant version of VISTA enhancer hs111
    for $DeepNeuralTube^o$ (top two panels) and $DeepNeuralTube^e$ (bottom two panels) and enhancer-reporter
    assay for reference and mutant version of hs111 (right two panels)\cite{
    viselVISTAEnhancerBrowsera2007, kosickiVISTAEnhancerBrowser2025}. 
    \textbf{h}, contribution score for
    $DeepNeuralTube^o$ (top) and $DeepNeuralTube^e$ (bottom) of a genomic human neural crest
    enhancer (chr1:161200798-161201298; hg38) and enhancer-reporter assay in chicken embryo (HH4; right). Green is enhancer signal, pink is
    constitutive enhancer that serves as electroporation control.
    Also see Figure S2
  }
\end{figure}

@@

Focusing on neural progenitors, we identified topics representing cells and regions of different dorso-ventral
identities in both the organoid and the embryo data (\textbf{Fig. 3a}). 
From ventral to dorsal this is topic 33, 38, 36, 54 and 48 for the organoids and
topic 34, 38, 79, 88 and 58 for the embryo. 
As an example we first investigated two enhancers of \textit{SHH} that are strongly conserved and specific to the 
floorplate: SFPE1 and SFPE2\cite{jeongDistinctRegulatorsShh2003}. 
\textit{SHH} is most strongly expressed in the cells of the most ventral topic (33 and 34 resp. for the organoids and the embryo; \textbf{Fig. 3b}).
SFPE1 is located upstream of \textit{SHH} and SFPE2 in its second intron. Both enhancers are indeed most accessible in the cells belonging to the ventral topic (\textbf{Fig. 3c}). 
When we tested the enhancer activity of SFPE1 in chicken embryos we could confirm its floor plate specific activity
(\textbf{Fig. 3d}).
Explaining the accessibility of both enhancers using DeepNeuralTube reveals two FOX (one partial) and two RFX (one partial) binding sites in SFPE1 and two FOX, a ZEB and a TEAD binding site in SFPE2 (\textbf{Fig. 3e}).\par

To investigate the enhancer code underlying dorso-ventral patterning of neural progenitors further, we clustered 
and aligned \textit{de novo} motifs, identified in dorso-ventral regions, based on pairwise motif similarities. 
Next, we calculated the average contribution for all identified motif instances per cluster and for each topic 
resulting in a dorso-ventral enhancer code-table (\textbf{Fig. 3f}).
Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns underlying the dorso-ventral states and the 
importance of the instances of these patterns is highly correlated across both models (0.75 average Pearson correlation; \textbf{Fig. 3f}). 
For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD and RFX instances 
are important (\textbf{Fig. 3f-g}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo (\textbf{Fig. 3h} and Fig. S3). 
In subsequent more dorsal neural progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3f-h)}); 
PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3f-h} and Fig. S3) and 
ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3f-h} and Fig. S3). 
For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), 
SOX instances are important correlating with the expression of \textit{SOX2/3} (\textbf{Fig. 3f-h} and Fig. S3). 
Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3f} and Fig. S3). 
Finally, we calculated the Jaccard index of genomic regions based on the presence of instances of each identified pattern to investigate which patterns often co-occur in the same region (\textbf{Fig. 3i}). 
FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns tend to be present in separate regions (\textbf{Fig. 3i}).\par

@@

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) 
    \textbf{b}, Boxplots of the expression of \textit{SHH} in organoid (top) and embryo (bottom) across different dorso-  ventral topics ($n_{topic 33}$ = 4,571, $n_{topic 38}$ = 4,521, $n_{topic 36}$ = 6,120, $n_{topic 54}$ = 2,808, and $n_{topic 48}$ = 5,390 for organoid cells;
      and $n_{topic 34}$ = 637, $n_{topic 38}$ = 1,093, $n_{topic 79}$ = 3,939, $n_{topic 88}$ = 1,861, and $n_{topic 58}$ = 1,773 for embryo cells).  With n representing the number of cells.
    \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr7:155,799,664-155,827,483 ; hg38. SFPE2 and SFPE1 are highlighted. 
    \textbf{d}, Enhancer reporter assay in chicken embryo. Blue is DAPI, pink is constitutive electroportion control,
    green is GFP driven by SFPE1 enhancer.
    \textbf{e}, Nucleotide contribution scores for SFPE2 (left) and SFPE1 (right) for organoid topic 33 (top) and embryo topic 34 (bottom). 
    \textbf{f}, Code table showing average contribution score of pattern instances for each dorso-ventral topic. Contribution scores for $DeepNeuralTube^o$ are not outlined. Contribution scores for $DeepNeuralTube^e$ are outlined and negated. 
    \textbf{g}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) for dorso-ventral topics. 
    \textbf{h}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) dorso-ventral topics. 
    \textbf{i}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. 
    \textbf{j}, Nucleotide contribution score for chr18:77367159-77367659; hg38 for organoid topic 33 (top) and embryo
    topic 34 (bottom).
    \textbf{k}, Prediction scores for chr18:77367159-77367659; hg38 for doros-ventral topics in organoid (top) and
    embryo (bottom).
    \textbf{l}, Enhancer reporter assay in chicken embryo. Blue is DAPI, pink is constitutive electroportion control,
    green is GFP driven by chr18:77367159-77367659; hg38 enhancer. Scale bar represents \SI{500}{\micro\meter}. 
    O. organoid. E. Embryo.
    Also see Figure S3 and S4.
}
\end{figure}

\subsection*{
  Distinct homeodomain transcription factor binding sites encode anterior-posterior patterning
}

@@

$DeepNeuralTube^e$ was able to distinguish the different AP domains but grouped topics of similar domains
% TODO: S.A.: How does the organoid model score all these AP topics?
together (i.e., Topic 59, 31 and 62; Topic 52 and 71), resulting in four distinct AP domains (\textbf{Fig.4 c-d}) and 
for this it made use of four distinct homeodomain TF binding motifs, one for each domain (\textbf{Fig. 4e}). 
From anterior to posterior this is \textit{TAATTAG}, \textit{TAATCCCT}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA}.
Notably, the patterns corresponding to the two posterior most topics correspond to homeobox dimer-motifs. 
Previously it had been hypothesized that genome-wide specific binding of homeodomain TFs is accomplished through
cooperative binding of two TFs\cite{mannChapter3Hox2009}. 
Indeed, the presence of instances of the \textit{TCATGWTGANTGA} motif correlates with the expression of both \textit{PAX5/8} and \textit{EN1/2} 
and the prescence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig. 4d-e} and Fig. S5). 
In contrast, patterns corresponding to the two anterior most topics correspond to monomer motifs, 
correlating with the expression of \textit{LHX2/9}, \textit{RAX} and \textit{SIX3/6} for the most anterior topic and 
with the expression of \textit{OTX1/2}, \textit{EMX2}, \textit{DMBX1} and \textit{LMX1A} for the second most anterior
topic (\textbf{Fig. 4d-e} and Fig. S5)\par
%TODO: I'm surprised to see EMX2 (forebrain) together with OTX1/2 and DMBX1 (midbrain)

@@

\afterpage{\clearpage}
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/main/Figure_4.png}
  \caption{
    \textbf{An unsophisticated enhancer code underlies anterior-posterior patterning in neural progenitors.} \\
  \textbf{a}, UMAP of embryo neural progenitors colored by cell-topic probabilities in RGB color scales for
anterior-posterior topics. 
  \textbf{b}, Heatmap showing pairwise Pearson correlation coefficients of region-topic probabilities for dorso-ventral and anterior-posterior topics. 
  \textbf{c}, Heatmap showing prediction scores of $DeepNeuralTube^e$ on the topic 3,000 regions per topic (y-axis) for each anterior-posterior topics (x-axis). 
  \textbf{d}, Heatmap showing the log of the number of seqlets for each identified pattern (y-axis) for each anterior-posterior topic (x-axis). 
  \textbf{e}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in embryo
  anterior-posterior topics. 
  \textbf{f}, Boxplot showing prediction score before (wt) and after (alt) substituting all instances of \textit{CTAATTAG} with \textit{CTAATCCCT} (top) or substituting all instances of \textit{CTAATCCCT} with \textit{CTAATTAG} (bottom) in the topic 3,000 regions of either Topic 61 or Topic 31. 
  \textbf{g}, Boxplot showing prediction scores for all anterior-posterior topics after embedding 0 to 5 instances of \textit{CTAATTAG} while optimizing their location for Topic 61 (top) and after embedding 0 to 5 instances of \textit{CTAATCCCT} while optimzing their location for Topic 31 (bottom); n = 200. 
  \textbf{h}, Boxplot showing prediction scores for all anterior-posterior topics after embedding 0 to 5 instances of \textit{CTAATTAG} while optimizing their location for Topic 31 (top) and after embedding 0 to 5 instances of \textit{CTAATCCCT} while optimzing their location for Topic 61 (bottom); n = 200. 
  \textbf{i}, receiver-operator curve (ROC) (top) and precision-recall curve (bottom) for logistic regression classifier classifying genomic regions as either Topic 61 or Topic 31 based on motif scores for \textit{CTAATTAG} and \textit{CTAATCCCT} motifs. 
  U.: UMAP. AP: anterior-posterior, DV: dorso-ventral, iter: iteration. 
  Also see Figure S5.
}
\end{figure}
\afterpage{\clearpage}

@@

\afterpage{\clearpage}
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/main/Figure_5.png}
  \caption{
    \textbf{Deciphering the enhancer code underlying pre-migratory and migratory neural crest and facial mesenchyme.} \\
  \textbf{a}, UMAP of organoid (top two) and embryo (bottom two) neural crest cells colored by cell-topic probabilities in RGB color scales for neural crest related topics. 
  \textbf{b}, Code table showing average contribution score of pattern instances for each neural crest related topic.  Contribution scores for $DeepNeuralTube^e$ are outlined and negated. 
  \textbf{c}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) of neural crest related topics. 
  \textbf{d}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) for neural crest related topics. 
  \textbf{e}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. 
  \textbf{f}, Manhattan plot showing -log p value of the SNP being associated with variations in facial shape\cite{whiteInsightsGeneticArchitecture2021} (positive y-axis) and the maximum (across classes of the model) of the absolute value of the difference in prediction score between the reference and alternative allele (negative y-axis; circles for $DeepNeuralTube^o$ and crosses for $DeepNeuralTube^e$) across the human genome. Lead SNPs are shown with a bigger dot size. 
  \textbf{g}, Zoom in of chromosome one with the locus of SNP rs1555067 shown as an inset, the SNP itself is shown in red. 
  \textbf{h}, Nucleotide contribution scores for Topic 58 of $DeepNeuralTube^o$ of the references (top) and alternative (bottom) allele of SNP rs1555067. The location of the SNP is indicated with a dotted line. 
  U: UMAP. O.: organoid. E.: Embryo. bps: base pairs. 
  Also see Figure S6 and S7.
  }
\end{figure}
\afterpage{\clearpage}

@@

For the cells representing early neural differentiation, we identified several topics for different stages of neural fate
acquisition in both organoid and embryo (\textbf{Fig. 6a}). Again we made use of $DeepNeuralTube^o$ and
$DeepNeuralTube^e$ to generate a neuron code-table (\textbf{Fig. 6b}). 
Both models found similar patterns and the importance of the instances of these patterns is correlated across
both models (0.47 average Pearson correlation coefficient; \textbf{Fig. 6b}).\par

@@
